Investigating the Impact of Macroeconomic Indicators on the FTSE 250 Index Using Machine Learning Methods

### 2024-08-15 tex/paper.tex
draft

A key task which economic policy makers and investors undertake is to 
understand and predict how economic changes influence stock market behaviour, 
which is critical for making informed policy and investment decisions. 
This modelling of the impact of macroeconomic indicators on stock markets is typically achieved by a variety of methods: statistical analysis, econometric modelling, scenario planning, and advanced machine learning techniques. However, it is fraught with challenges that require careful consideration, advanced techniques, and a deep understanding of both economics and financial markets. In particular, there is a challenge to find a subset of macroeconomic indicators which can most accurately capture stock market dynamics. 
In addition, macroeconomic indicator data is of such low-granularity that it precludes the use of certain deep learning neural network methods for prediction purposes.

This study aims to explore the relationship between macroeconomic indicators and stock markets 
through modelling. Its primary objective is to address questions such as how specific 
macroeconomic indicators influence stock market prices in major economies and whether machine 
learning models can substitute traditional econometric models in predicting stock market prices. 
Ultimately, it aims to investigate three aspects of modelling the impact of macroeconomic 
indicators on stock markets: 
\begin{enumerate}
    \item To investigate how macroeconomic indicators influence the FTSE 250 index
    \item To investigate the short-term and long-term effects of macroeconomic variables on the price of the FTSE 250 index
    \item To investigate whether machine learning techniques can provide accurate short-term and long-term forecasts of the price of the FTSE 250 index when using a low-granularity dataset.

@@

In order to analyse the dynamic relationship between variables in time series data, this study will employ the Autoregressive Distributed Lag (ARDL) model in conjunction with an Error Correction Model (ECM). By integrating the ARDL model with the ECM, a comprehensive understanding of both the immediate and longer-term relationships between variables is obtained, which is valuable for robust economic analysis and policy-making.
This research also employs radial basis function neural networks to predict the price of the FTSE 250 index. This artificial deep learning network is particularly useful for tasks like function approximation, pattern recognition, and regression when the dataset is of low-granularity.

@@

In the short-term, which defined as 5 months, 
as per the optimum lag for the FTSE250 in the ARDL model, the market movements
are more closely tied to recent deviations in the economy, because theoretically
the short-run is not exposed to the various unforeseen external shocks and structural changes
that dilute the relationship between these variables and the stock market in the 
long-run. Essentially, as the prediction window expands, 
more uncertainty and unknowns are introduced, which translates into 
lower predictive quality. Furthermore, as the time horizon extends into the long-run, fundamental factors 
begin to matter a lot more. The compound effect of earnings, cash flows and
managerial efficiencies begins to outweigh the 
influence of fluctuating macroeconomic sentiment. Thus, in the short-run 
a different phenomenon is observed.

@@

Further research should endeavour to include other macroeconomic variables such as 
foreign direct and portfolio investment, commodity prices and economic growth.
An autoregresive model, falling outside the immediate scope of the current study, could also be tried, including the past values of the 
FTSE250 in the input space to provide a more accurate model. In addition, 
a more granular data set could be utilised by applying a 
Brownian bridge algorithm to generate daily macroeconomic data and 
subsequently testing the RBF-NN against other neural networks 
such as auto-regressive LSTMs, as well as other machine learning methods suited for 
forecasting such as Gradient Boosting Machines and Support Vector Machines.

>>>>

A key task which economic policy makers and investors undertake is to 
understand and predict how economic changes influence stock market behaviour, 
which is critical for making informed policy and investment decisions. 
This modelling of the impact of macroeconomic indicators on stock markets is typically achieved by a variety of methods: statistical analysis, econometric modelling, scenario planning, and advanced machine learning techniques. However, it is fraught with challenges that require careful consideration, advanced techniques, and a deep understanding of both economics and financial markets. In particular, there is a challenge to find a subset of macroeconomic indicators which can most accurately capture stock market dynamics. 
In addition, macroeconomic indicator data is of such low granularity that it precludes the use of certain deep learning neural network methods for prediction purposes.

This study aims to explore the relationship between macroeconomic indicators and stock markets 
through modelling. Its primary objective is to address questions such as how specific 
macroeconomic indicators influence stock market prices in major economies and whether machine 
learning models can substitute traditional econometric models in predicting stock market prices. 
Ultimately, it aims to investigate three aspects of modelling the impact of macroeconomic 
indicators on stock markets: 
\begin{enumerate}
    \item To investigate how macroeconomic indicators influence the FTSE 250 index
    \item To investigate the short-term and long-term effects of macroeconomic variables on the price of the FTSE 250 index
    \item To investigate whether machine learning techniques can provide accurate short-term and long-term forecasts of the price of the FTSE 250 index when using a coarse-grain dataset.

@@

In order to analyse the dynamic relationship between variables in time series data, this study will employ the Autoregressive Distributed Lag (ARDL) model in conjunction with an Error Correction Model (ECM). By integrating the ARDL model with the ECM, a comprehensive understanding of both the immediate and longer-term relationships between variables is obtained, which is valuable for robust economic analysis and policy-making.
This research also employs radial basis function neural networks to predict the price of the FTSE 250 index. This artificial deep learning network is particularly useful for tasks like function approximation, pattern recognition, and regression when the dataset is of low granularity.

@@

In the short-term the market movements
are more closely tied to recent deviations in the economy. This is because,
theoretically, 
the short-run is not exposed to the various unforeseen external shocks and structural changes
that dilute the relationship between these variables and the stock market in the 
long-run. As the prediction window expands, 
more uncertainty and unknowns are introduced, which translates into 
lower predictive quality. Furthermore, as the time horizon extends into the long-run, fundamental factors 
begin to matter a lot more. The compound effect of earnings, cash flows,
innovations and managerial efficiencies begins to outweigh the 
influence of fluctuating macroeconomic variables. Essentially, firms become accustomed 
to the changed economic environment and over time. Thus, in the short-run 
a different phenomenon is observed.

@@

Further research should endeavour to include other macroeconomic variables such as 
foreign direct and portfolio investment, commodity prices and economic growth.
An autoregressive neural network could also be tried, including the past values of the 
FTSE250 in the input space to provide a more accurate model. In addition, 
a more granular data set could be utilised by applying a 
Brownian bridge algorithm to generate daily macroeconomic data. This would allow 
the testing of the RBF-NN against other neural networks 
such as the LSTM. Other machine learning methods suited for 
forecasting such as Gradient Boosting Machines and Support Vector Machines could 
also be investigated in future work in this problem domain.
